Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-29 relwork.tex
polish relwork, conclusion

\add{
Kolanski and Klein verify the critical step of updating an already-located L1 entry to map a new page
(ARM assembly corresponding roughly to Lines \ref{line:l1entry_storeC} and \ref{line:l1entry_setpresent}
in our Figure \ref{fig:mapping_codeC}), but ignore the essential code
preceding that step --- which as our \lstinline|walkpgdir| and \lstinline|pte_get_next_table| verifications
demonstrate, side-steps a significant amount of complexity and critical reasoning tasks.
We have verified the entirety of the software page table walk up to mapping a new page, aside from a trusted physical memory allocator resembling
\lstinline|malloc|~\cite{Chlipala2013Bedrock,wickerson2010explicit}.
}
 As a consequence of tackling this larger verification challenge, we are also the first to formally specify large
 segments of the self-referential portion of an OS kernel's virtual memory management invariants (per Sections \ref{sec:p2vC} and \ref{sec:selfconditional}),
 and to reason about converting back from physical addresses to virtual addresses efficiently.
}
\looseness=-1

@@

\replace{Unlike our work, Kolanski and Klein prove very useful embedding theorems stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts).}{
 Kolanski and Klein prove that updates to any memory locations that are not part of the page tables
 do not affect the interpretation of other memory addresses.
 This implies that programs that do not modify memory mappings could be reasoned about without
 concern for mappings.
}
While we have not \add{formally} proven such a result \add[of our machine model, it holds (just as for real hardware)],
an analagous result \replace{{should} hold}{holds} of our work: consider that the doubles for the \texttt{mov} instructions
that access memory behave just as one would expect for a VM-ignorant logic~\cite{Chlipala2013Bedrock}.
With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle be extended to account for pageable points-to assertions by adding \del{additional}
disjunctions to an extended points-to definition; embedding ``regular'' separation logic into such a variant
is the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.

\add{
 No prior work has dealt with combined address space changes and context switching. \textsc{XCAP}~\cite{ni2007contexts} deals
 only with usermode threading (in a single process). \textsc{CertiKOS} and \textsc{seL4} trust assembly primitives for
 context switches, in addition to not modeling address translation for executing code.
}
% \todo[inline]{author response talks about syeda and klein here briefly, and it fits, but we also want the longer discussion}

@@

The only other attempt to verify a change of address space was by \citet{syeda2018program,syeda2020formal}, whose focus was on TLB maintenance (which we
do not model). However, they verified only the 4 instructions to switch address spaces and update the TLB on an ARM processor, in isolation
(i.e., not the full context switch including changing stacks with address spaces).
The specification they proved for those instructions did not address program invariants that may be valid in one address space and not the other,
so is not flexible enough to extend directly to a full context switching primitive as in Figure \ref{fig:swtchC}\replace{,}{.}
\del{in addition to inheriting other limitations from {the work they extend, discussed above}.}
\add{Their logic adds an assertion tracking known-inconsistent addresses (i.e., recently-unmapped by a table update or change in page table root)
and their memory access rules require accessed memory to lie outside that set.
However their logic is proven sound against a generalized semantics where any page table change adds \emph{all}
addresses to the inconsistent set, which is too imprecise for finer-grained unmapping.
}
The right \replace{eventual}{general} solution would be to combine our work
and \add{an extension of} theirs, \add{a substantial project} which we leave to future work\replace{,}{.}\del{ as combining both the models and reasoning principles would be highly non-trivial.}

>>>>

\add{
Kolanski and Klein verify the critical step of updating an already-located L1 entry to map a new page
(ARM assembly corresponding roughly to Lines \ref{line:l1entry_storeC} and \ref{line:l1entry_setpresent}
in our Figure \ref{fig:mapping_codeC}), but ignore the essential code
preceding that step --- which as our \lstinline|walkpgdir| and \lstinline|pte_get_next_table| verifications
demonstrate, side-steps a significant amount of complexity and critical reasoning tasks.
We have verified the entirety of the software page table walk up to mapping a new page, aside from a trusted physical memory allocator resembling
\lstinline|malloc|~\cite{Chlipala2013Bedrock,wickerson2010explicit}.
 As a consequence of tackling this larger verification challenge, our work is the first to formally specify large
 segments of the self-referential portion of an OS kernel's virtual memory management invariants (per Sections \ref{sec:p2vC} and \ref{sec:selfconditional}),
 and to reason about converting from physical addresses to virtual addresses efficiently.
\looseness=-1
}

@@

\replace{Unlike our work, Kolanski and Klein prove very useful embedding theorems stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts).
}{
 Kolanski and Klein prove that updates to any memory locations that are not part of the page tables
 do not affect the interpretation of other memory addresses, just like on real hardware.
 This implies that programs that do not modify memory mappings could be reasoned about without
 concern for mappings.
}
\replace{While we have not proven {such a result},
an analagous result \replace{{should} hold}{holds} of our work:}{
 An analagous result should holds of our model (though we have not proven it).
}
\replace{ consider that the doubles for the}{Informally this is visible in the rules for}
\texttt{mov} instructions,
\replace{that access memory behave just as one would expect for}{which are nearly identical to rules in a} VM-ignorant logic~\cite{Chlipala2013Bedrock,ni2007contexts}.
\replace{With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle}{
  In principle both our approach and Kolanski and Klein's could
 } be extended to account for pageable points-to assertions by adding \del{additional}
disjunctions to an extended points-to definition\add{,}\del{; embedding ``regular'' separation logic into such a variant
is} \add{which would be}
 the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.
\looseness=-1

% \todo[inline]{author response talks about syeda and klein here briefly, and it fits, but we also want the longer discussion}

@@

The only \replace{other attempt to verify a change of address space was by}{work to tackle TLB code verification was}
\citet{syeda2018program,syeda2020formal},
\replace{whose focus was on TLB maintenance (which we do not model)}{who are also the only prior work on verifying address-space-aware context switching}.
However, they verified only the 4 instructions to switch address spaces and update the TLB on an ARM processor, in isolation
(i.e., not the full context switch including changing stacks with address spaces).
The specification they proved for those instructions did not address program invariants that may be valid in one address space and not the other,
so is not flexible enough to extend directly to a full context switching primitive as in Figure \ref{fig:swtchC}\replace{,}{.}
\del{in addition to inheriting other limitations from {the work they extend, discussed above}.}
\add{Their logic adds an assertion tracking known-inconsistent addresses (i.e., recently-unmapped by a table update or change in page table root)
and their memory access rules require accessed memory to lie outside that set.
However their logic is proven sound against a generalized semantics where any page table change adds \emph{all}
addresses to the inconsistent set, which is too imprecise for finer-grained unmapping.
}
The right \replace{eventual}{general} solution would be to combine our work
and \add{an extension of} theirs, \add{a substantial project} which we leave to future work\replace{,}{.}\del{ as combining both the models and reasoning principles would be highly non-trivial.}
\add{
 No other prior work has considered address space changes during context switching. \textsc{XCAP}~\cite{ni2007contexts}
 and Bedrock~\cite{Chlipala2011Bedrock,Chlipala2013Bedrock,Chlipala2015webapp} deal
 only with usermode threading (in a single process). \textsc{CertiKOS} and \textsc{seL4} trust assembly primitives for
 context switches, and do  not model address translation for executing code.
 \looseness=-1
}
